Letter: Performance models for \\ high-resolution ice sheet simulations

### 2022-06-23 paper/perfmod.tex
edits to mostly complete introduction

@@

\abstract{\today. FIXME}

@@

\subsection{Introduction}

FIXME INCREASING UNDERSTANDING OF TIME STEP LIMITS \citep{Chengetal2017,Robinsonetal2022}; MORE THAN SIA CASE \citep[for comparison]{HindmarshPayne1996}, POSSIBILITY OF IMPLICIT SOLVERS \citep{Bueler2016,WirbelJarosch2020}, NEED PERFORMANCE MODEL TO UNDERSTAND TRADE-OFFS

FIXME \citep{GreveBlatter2009,SchoofHewitt2013} for SIA and Stokes equations

FIXME We expect that time-stepping glacier and ice-sheet numerical models will become faster, and thus more useful tools for glaciologists, when multigrid is applied to the solution of the problem(s) solved at each time-step.  This hope can be understood and assessed by a \emph{performance model} for the scheme, one which combines computational work estimates for subproblems into an asymptotic, in the limit of highly-refined meshes, description of of the work required for a given simulation task.

\subsection{Performance models}

@@

\subsection{Discussion and Conclusion}

>>>>

\newcommand{\Divx}{\nabla_\bx \cdot}
\newcommand{\gradx}{\nabla_\bx}

@@

\abstract{High-resolution numerical ice sheet models, with horizontal resolutions of a few kilometers or less, compute evolving ice geometry and velocity fields using various stress-balance approximations and boundary conditions.  These models usually require many short, stability-limited time steps because ice thickness is updated following each velocity solution.  Practical performance in the high-spatial-resolution limit is degraded by the stability conditions of such explicit time-stepping.  In this short note, which considers the shallow ice approximation and Stokes models as end members, we attempt to clarify numerical model performance by quantifying the simulation cost per model year in terms of the number of degrees of freedom.  The performance of coming generations of implicit time-stepping numerical models is then assessed.}

@@

\section{Introduction}

Numerical ice sheet (glacier) models with evolving ice geometry are now in common use for scientific questions such as quantification of future sea level rise from changes in the Antarctic \citep{Seroussietal2020} and Greenland \citep{Goelzeretal2020} ice sheets, interpretation of the paleoglacial record \citep{Weberetal2021}, and quantification of long-term glacial erosion \citep{SeguinotDelaney2021}, among other applications.  It is generally accepted that horizontal mesh (grid) cells must be smaller than about 10 km in order to generate valid results.  Whether using local mesh refinement \citep[for example]{Hoffmanetal2018} or not, resolutions of a few kilometers to one kilometer \citep{SeguinotDelaney2021} or less \citep{Aschwandenetal2019} are increasingly used at ice sheet scale.  With currently-available computational resources, sub-kilometer meshes should become common for long-duration, whole ice sheet studies.

Current-generation ice sheet models use a variety of stress balances, from simple approximations like the shallow ice approximation (SIA), through ``hybrid'' \citep{Robinsonetal2022,Winkelmannetal2011} and higher-order balances, to the non-shallow Stokes approximation.  With very few exceptions, however, current models alternate between solving the stress balance for velocity, over the geometry determined by the previous time step, and then updating the geometry, i.e.~the ice thickness and/or surface elevation.  The geometry-update operation, using the mass continuity or surface kinematical equations \citep{GreveBlatter2009}, is where interaction between the ice sheet and the climate is (primarily) simulated, especially through surface mass balance, sub-shelf (basal) mass balance, and calving processes.  Such schemes are using \emph{explicit} time-stepping for the coupled mass and momentum system which describes ice dynamics.

For simpler partial differential equation problems, the conditional stability of explicit time-stepping schemes is well-understood \citep{LeVeque2007}.  The stability limitations of explicit SIA models have also been understood for some time \citep{HindmarshPayne1996}, but recent studies have focussed on the time-step limits of hybrid, higher-order, and Stokes dynamics models \citep{Chengetal2017,Robinsonetal2022}, or on lengthening the steps \citep{LofgrenAhlkronaHelanow2021}.

However, \emph{implicit} time-stepping should also be considered.  Here the velocity and geometry (thickness or surface elevation) are updated simultaneously by solving a coupled mass-momentum problem.  While implicit time-stepping is a routine solution process for certain simpler partial differential equations \citep[for example]{LeVeque2007}, for ice sheets the implicit step must simultaneously compute the velocity and the \emph{domain on which the velocity is defined}.  The problem is of free-boundary type in map-plane (horizontal) directions \citep{SchoofHewitt2013}, as well as in the easily-resolved vertical direction.  An implicit strategy has been demonstrated at high-resolution in the simplest frozen-base, isothermal SIA case \citep{Bueler2016}.  The same reference shows how the steady-state problem \citep{JouvetBueler2012} can often be solved; steady-state solutions are an implicit time step of infinite duration.  Early work applying a semi-implicit time step using Stokes dynamics \citep{WirbelJarosch2020} illuminates some of the techniques, and difficulties, needed to make such a free-boundary problem work with a membrane-stress-resolving balance.

This is the context in which we will relate time-stepping and stress-balance choices to overall computational effort.  The reader is assumed to have access to the SIA and Stokes stress-balance equations \citep{GreveBlatter2009,SchoofHewitt2013}, regarded here as end members of current-usage stress-balance approximations \citep{Robinsonetal2022}, and familiarity with the mass continuity and surface kinematical equations \citep{GreveBlatter2009}.

However, the manner of presenting the mass continuity equation should first be examined.  For an incompressible ice sheet with thickness $H(t,\bx)$, vertically-averaged velocity $\bu(t,\bx)$, and climatic-basal mass balance $a(t,\bx)$, this equation says
\begin{equation}
\frac{\partial H}{\partial t} + \Divx \left(\bu H\right) = a, \label{eq:masscontinuity}
\end{equation}
where $\bx=(x,y)$ denotes horizontal coordinates.

Equation \eqref{eq:masscontinuity} suggests that ice sheets change geometry in an essentially advective manner, but this appearance is deceiving, or at least over-simplified, regarding the growth of numerical instabilities.  This is because ice flows dominantly downhill.  Thus, as addressed by linearized analysis \citep{Chengetal2017,Robinsonetal2022}, when thickness perturbations grow unstably under explicit time-stepping, i.e.~with too large a step, they do so by a mix of advective and diffusive mechanisms.

Let $s(t,\bx)=H(t,\bx)+b(\bx)$ denote the surface elevation for bed elevation $b(\bx)$.  If the numerical thickness is perturbed then the velocity $\bu$ often responds by growing in the downhill direction $-\gradx s$, a direction often correlated to $-\gradx H$.  In membrane-stress-resolving models this happens through the nonlocal solution of the stress balance; the driving stress is along $\gradx s$.  Equation \eqref{eq:masscontinuity} has velocity $\bu$ which is really a non-local function ``$\bu(H,\grad_x s)$''; a solution of the stress balance equations is required to evaluate this function.

Thus the mass continuity equation is substantially, though not wholely, diffusive.  The diffusive nature dominates in the small-parameter limit which generates the SIA, in which a revised form of the mass continuity equation holds:
\begin{equation}
\frac{\partial H}{\partial t} = \Divx \left(D \gradx s \right) + a  \label{eq:siamasscontinuity}
\end{equation}
where
\begin{equation}
D = \Gamma H^{n+2} |\gradx s|^{n-1} \label{eq:siadiffusivity}
\end{equation}
is the nonlinear diffusivity.  Here $\Gamma = 2 A (\rho g)^n/(n+2)$ is a positive constant in the isothermal case, with $A$ the ice softness, $\rho$ the ice density, $g$ gravity, and $n$ the Glen exponent in the flow law \citep{GreveBlatter2009}.

Unlike equation \eqref{eq:masscontinuity}, equation \eqref{eq:siamasscontinuity} does not hold for all models, but the essentially-geometric diffusivity $D$ can be computed from equation \eqref{eq:siadiffusivity}, regardless.  In any numerical model of grounded ice sheets, large values of $D$ will coincide with locations of unstable mode growth if explicit time-steps are chosen too large.

FIXME no characteristics

FIXME SITUATION COMPLICATED AND PRONE TO MISUNDERSTANDING; NEED PERFORMANCE MODEL TO UNDERSTAND TRADE-OFFS; We expect that time-stepping glacier and ice-sheet numerical models will become faster, and thus more useful tools for glaciologists, when multigrid is applied to the solution of the problem(s) solved at each time-step.  This hope can be understood and assessed by a \emph{performance model} for the scheme, one which combines computational work estimates for subproblems into an asymptotic, in the limit of highly-refined meshes, description of of the work required for a given simulation task.

\section{Performance models}

@@

\section{Discussion and Conclusion}
